---
title: "Stroke Dataset"
subtitle: " Building and evaluating a prediction model"
author: "Sivakamakshi Muthu Kumarasamy, ID = 25240080"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: kable
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(table1)
library(dplyr)
library(plotly)
library(caret)
library(pROC)
library(randomForest)
library(gbm)
library(PRROC)
library(dcurves)
library(patchwork)
library(kableExtra)
library(scales)
library(MASS)
library(rpart)
library(rpart.plot)
library(pROC)
library(e1071)
library(tree)
library(BART)

```

## Prediction Goal


## Dataset

Dataset: Stroke Prediction Dataset

Source: https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset?resource=download

The study design is cross-sectional with adult patients with demographic, clinical and lifestyle characteristics.


## Data loading and Pre-processing

```{r}

set.seed(123)
stroke <- read.csv("stroke dataset.csv",
                   stringsAsFactors = FALSE,
                   na.strings = c("NA", "N/A", "", " "))

stroke <- stroke %>%
  filter(gender != "Other") %>%
  dplyr::select(-id) %>%
  mutate(
    bmi            = as.numeric(ifelse(bmi == "N/A", NA, bmi)),
    bmi            = ifelse(is.na(bmi), median(bmi, na.rm = TRUE), bmi),
    stroke         = factor(stroke, levels = c(0, 1), labels = c("No", "Yes")),
    gender         = factor(gender),
    ever_married   = factor(ever_married),
    work_type      = factor(work_type),
    Residence_type = factor(Residence_type),
    smoking_status = factor(smoking_status),
    hypertension   = factor(hypertension),
    heart_disease  = factor(heart_disease)
  )
```

```{r}
names(stroke)
```
```{r}
summary(stroke)
```

```{r}
colSums(is.na(stroke))
```

```{r}
n_stroke <- sum(stroke$stroke == "Yes")
n_stroke
```

```{r}
male <- sum(stroke$gender == "Male")
female <- sum(stroke$gender == "Female")

male
female
```
## Exploratory Visualisation

```{r}
#Age distribution by Stroke status

age_plot <- ggplot(stroke, aes(x = age, fill = stroke)) +
  geom_histogram(bins = 40, alpha = 0.75, position = "identity") +
  labs(title = "Age distribution by Stroke status", x = "Age", y = "Count", fill = "Stroke") + 
  theme_minimal()

# Stroke status by average glucose levels
glucose_plot <- ggplot(stroke, aes(x = avg_glucose_level, fill = stroke)) +
  geom_histogram(bins = 40, alpha = 0.75, position = "identity") +
  labs(title = "Stroke status by average glucose levels", x = "Average glucose levels", y = "Count", fill = "Stroke") + 
  theme_minimal()


# Stroke rate by BMI category
bmi_plot <- ggplot(stroke, aes(x = stroke, y = bmi, fill  = stroke)) +
  geom_boxplot() +
  labs(title = "Stroke status by BMI levels", x = "Stroke", y = "BMI", fill = "Stroke") +
  theme_minimal()

#Stroke rate by heart disease

heart_plot <- stroke %>%
  count(heart_disease, stroke) %>%
  group_by(heart_disease) %>%
  mutate(proportion = n / sum(n)) %>%
  filter(stroke == "Yes") %>%
  ggplot(aes(x = heart_disease, y = proportion, fill = heart_disease)) +
  geom_col() +
  scale_y_continuous(labels = percent_format())+
  labs(title  = "Stroke status by heart diseases", x = "Heart Disease  0 = No, 1 = Yes", y = "Stroke", fill = "Stroke") +
  theme_minimal()

#Stroke status by hypertension
  
htn_plot <- stroke%>%
  count(hypertension, stroke) %>%
  group_by(hypertension) %>%
  mutate(proportion = n / sum(n)) %>%
  filter(stroke == "Yes") %>%
  ggplot(aes(x = hypertension, y = proportion, fill = hypertension)) +
  geom_col()+
  scale_y_continuous(labels = percent_format())+
  labs(title = "Stroke status by hypertension", x = "Hypertension 0 = No, 1 = Yes", y = "Stroke", fill = "Stroke")+
  theme_minimal()

(age_plot | glucose_plot) /  bmi_plot / (heart_plot | htn_plot)
```

## Cross-validation

```{r}
# Considering 70% trainig data

set.seed(123)
n           <- nrow(stroke)
train_index <- sample(1:n, size = 0.7 * n)
train       <- stroke[train_index, ]
test        <- stroke[-train_index, ]

cat("Training set size:", nrow(train), "\n")
cat("Test set size:",     nrow(test),  "\n")

test_actual <- factor(test$stroke, levels = c("No", "Yes"))

cat("Training set stroke prevalence:", round(mean(train$stroke == "Yes")*100, 2), "%\n")
cat("Test set stroke prevalence:", round(mean(test$stroke == "Yes")*100, 2), "%\n")
```



## Logistic Regression model using only the observations of training set

```{r}
log_model <- glm(stroke ~., data = train, family = binomial)

summary(log_model)

```

```{r}
# Prediction

log_pred <- predict(log_model, newdata = test, type = "response")
log_pred_class <- factor(ifelse(log_pred > 0.5, "Yes", "No"), levels = c("No", "Yes"))
test_actual <- factor(test$stroke, levels = c("No", "Yes"))


confusionMatrix(log_pred_class, test_actual, positive = "Yes")
roc_curve <- roc(test$stroke, log_pred)
plot(roc_curve, col = "steelblue", main = "ROC Curve - Logistic Regression")
auc_full <- cat("AUC (logistic):", round(auc(roc_curve), 3), "\n")

```

```{r}
# Variable Selection

step_model <- stepAIC(log_model, direction = "both",trace = FALSE)
summary(step_model)
```

```{r}
AIC(log_model, step_model)
```

```{r}
# Predictions
step_pred <- predict(step_model, newdata = test, type = "response")
step_pred_class <- factor(ifelse(step_pred > 0.5, "Yes", "No"), levels = c("No", "Yes"))
test_actual <- factor(test$stroke, levels = c("No", "Yes"))

# Confusion Matrix
confusionMatrix(step_pred_class, test_actual, positive = "Yes")

# ROC and AUC
roc_step <- roc(ifelse(test_actual=="Yes",1,0), step_pred)
plot(roc_step, col="steelblue", main="ROC Curve - Stepwise Logistic Regression")
auc_step <- cat("AUC (stepwise):", round(auc(roc_step), 3), "\n")
```

```{r}

step_varimp <- varImp(step_model, scale = TRUE)
step_varimp_df <- as.data.frame(step_varimp)
step_varimp_df$Variable <- rownames(step_varimp_df)
rownames(step_varimp_df) <- NULL

imp_col <- names(step_varimp_df)[1]
step_varimp_df <- step_varimp_df %>% arrange(desc(.data[[imp_col]]))

step_varimp_df %>% kable(caption = "Variable Importance - Stepwise Logistic Regression")


ggplot(step_varimp_df, aes(x = reorder(Variable, .data[[imp_col]]), y = .data[[imp_col]])) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance - Stepwise Logistic Regression",
       x = "Variable",
       y = "Importance (scaled)") +
  theme_minimal()
```

```{r}
set.seed(123)
tree_model <- rpart(
  stroke ~ .,
  data    = train,
  method  = "class",
  control = rpart.control(cp = 0.0001, minsplit = 5, minbucket = 2)
)
printcp(tree_model)
plotcp(tree_model)

```
```{r}
set.seed(123)
rf_model <- randomForest(
  stroke ~ .,
  data       = train,
  ntree      = 200,
  importance = TRUE
)

print(rf_model)

# Predictions
rf_pred <- predict(rf_model, newdata = test)
rf_prob <- predict(rf_model, newdata = test, type = "prob")[,2]

# Confusion Matrix
confusionMatrix(rf_pred, test$stroke, positive = "Yes")
confusionMatrix(factor(log_pred_class), test$stroke, positive = "Yes")

# ROC and AUC
roc_rf <- roc(ifelse(test$stroke=="Yes",1,0), rf_prob)
plot(roc_rf, col="darkgreen", main="ROC Curve - Random Forest")
auc_rf <- auc(roc_rf)
cat("AUC for Random Forest:", round(auc_rf,3), "\n")


varImpPlot(rf_model, main="Variable Importance - Random Forest")

```

```{r}
best_cp     <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
tree_pruned <- prune(tree_model, cp = best_cp)

```
```{r}
tree_prob <- predict(tree_model, test, type = "prob")[, "Yes"]

tree_class <- factor(ifelse(tree_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))

confusionMatrix(tree_class, test_actual, positive = "Yes")


roc_tree <- roc(test$stroke, tree_prob)
auc_tree <- cat("AUC:", round(auc(roc_tree), 3), "\n") 
plot(roc_tree, col = "darkorange", main = "ROC – Decision Tree")

```

```{r}
plot(roc_curve,  col = "steelblue",   lwd = 2,
     main = "ROC comparison – all models")
plot(roc_step, col = "dodgerblue",  lwd = 2, lty = 2, add = TRUE)
plot(roc_tree, col = "darkorange",  lwd = 2, add = TRUE)
plot(roc_rf,   col = "forestgreen", lwd = 2, add = TRUE)
abline(a = 0, b = 1, lty = 3, col = "grey60")

legend("bottomright", bty = "n",
       legend = c(
         paste("Logistic (full)     AUC =", round(auc(roc_curve),  3)),
         paste("Logistic (stepwise) AUC =", round(auc(roc_step), 3)),
         paste("Decision tree       AUC =", round(auc(roc_tree), 3)),
         paste("Random forest       AUC =", round(auc(roc_rf),   3))
         ),
       col = c("steelblue","dodgerblue","darkorange","forestgreen","purple"),
       lwd = 2)

```
```{r}
model_summary <- data.frame(
  Model = c("Logistic Full", "Logistic Stepwise", "Decision Tree", "Random Forest"),
  AUC = c(auc_full, auc_step, auc_tree, auc_rf)
)
print(model_summary)
```

#Complete Tree Analysis

```{r}
table(stroke$stroke)
set.seed(123)
```


```{r}
tree.stroke <- tree(stroke ~., data = stroke)

tree.stroke.train <- tree(stroke ~., data = train)

print("Tree Summary")
summary(tree.stroke.train)

plot(tree.stroke.train)
text(tree.stroke.train, pretty = 0)


#Predictions

tree.pred <- predict(tree.stroke.train, test, type = "class")
table(tree.pred, test$stroke)
mean(tree.pred != test$stroke)
```


```{r}
#CV pruning



```